---
project:
  type: website
  output-dir: docs
website:
  title: "Ericka Quintero's ePortfolio"
  author: "Ericka Quintero"
  date: "15 May 2025"
  navbar:
    left:
      - href: index.qmd
        text: "Home"
      - href: about.qmd
        text: "About Me"
      - href: ericka_quintero_portfolio/ericka_quintero_portfolio.qmd
        text: "Queensland, Australia Project"
format:
  html:
    theme: cosmo
    css: styles.css
    toc: true
    echo: false
    warning: false
    message: false

editor: visual
---

# Project: Wave Activity in Queensland, Australia

## Data Description

Queensland monitors wave activity along its coastline using a network of floating buoys that measure wave height, movement, and sea surface conditions. These buoys, stationed at locations such as Cairns in the north and Gold Coast in the south, provide hourly updates that contribute to beach safety, maritime operations, and storm preparedness. While both northern and southern areas experience waves, the Gold Coast and surrounding regions typically see more powerful and consistent swells due to stronger Pacific Ocean currents. Northern waters, particularly near Cairns, tend to be calmer except during cyclones or intense monsoonal weather, when conditions can shift dramatically. This valuable data is sourced from Queensland's official wave monitoring program, available at: <https://www.qld.gov.au/environment/coasts-waterways/beach/monitoring>

## Data Cleaning

To get the data ready for analysis, I first ensured the date formal was consistent and extracted only the date from the full time stamp for easier grouping. I also organized the data by date and location, summarizing wave statistics such as average wave height, peak wave period, and maximum wave height. I filtered out only two locations to represent the north and the south, keeping only data from the Gold Coast and Cairns site. To make the site names easier, I simplified them by removing extra labels. Lastly, I arranged the data by location to make it more structured and ready for further analysis.

```{r}
#| label: data setup
library(tidyverse)
library(lubridate)
wavedata <- read_csv("dataraw/wave-7dayopdata.csv", skip = 1)
```

```{r}
#| label: data-cleaning

queensland_wave <- wavedata %>%
  mutate(DateTime = lubridate::ymd_hms(format(DateTime, format = "%Y-%m-%d %T %Z")),
         date = date(DateTime)) %>% 
  group_by(date, Site) %>% 
  summarize(
    avg_wave = mean(Hsig, na.rm = TRUE),
    peakwave = mean(Tp, na.rm = TRUE), 
    max_wave = mean(Hmax, na.rm = TRUE),
    SST = mean(SST, na.rm = TRUE)) %>%
filter(Site %in% c("Gold Coast Mk4", "Cairns Mk4")) %>%
  mutate(Site = recode(Site,
               "Gold Coast Mk4" = "Gold Coast",
               "Cairns Mk4" = "Cairns")) %>%
  arrange(Site)

write_csv(queensland_wave, "dataclean/queensland_wave")
```

## Visualization One

```{r}
#| label: visualization one - wave height
#| echo: false
#| fig-cap: "Signficant Wave Height in Queensland, Australia"
wave_height_plot <- queensland_wave %>% 
  pivot_longer(cols = contains("wave"),
               names_to = "wave_height",
               values_to = "wave_calc") %>%
  filter(wave_height != "peakwave") %>%
  ggplot(aes(x= date,
             y= wave_calc,
             group= wave_height,
             color= wave_height)) +
  geom_point() +
 labs(title= "Significant Wave Height in Queensland, Australia (2025)",
      x = "DateTime",
      y = "Wave Height (m)") +
  facet_wrap(~Site)
print(wave_height_plot)
```

This visualization shows how wave height changed over several days along the coast of Queensland Australia. Each dot represents the height of waves and the tallest waves recorded for specific days this year. Most waves measured around one to one and a half meters. However, the largest waves often exceeded three meters. These changes in wave height suggest that the ocean conditions varied from day to day, possibly due to shifts in weather or ocean activity.

## Visualization Two

```{r}
#| label: sea surface temperature
#| echo: false
#| fig-cap: "Sea Surface Temperature in Queensland, Australia"

wave_sst_plot <- queensland_wave %>%
  filter(Site %in% c("Gold Coast", "Cairns")) %>%
  ggplot(aes(x = Site, y = SST)) +
  geom_boxplot(fill = "lemonchiffon",
               color = "forestgreen") +
  labs(title = "Sea Surface Temperature in Northern and Southern Queensland Regions",
       x = "Site",
       y = "SST (°C)") +
  theme_light() +
  coord_flip()

print(wave_sst_plot)
```

This visualization highlights the difference in typical sea surface temperatures observed between Cairns in northern Queensland and the Gold Coast in the south.
